---
title: "Exercise: Modelling Binary Outcomes"
subtitle: "Session 3 · Quantitative Data Analysis"
author: "Your name here"
date: "2026-04-30"
format:
  html:
    toc: true
    number-sections: true
    code-tools: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
**Session business question:** *Can we predict which firms are likely to exit the market in the next year?*

Work through the tasks below. For each task, write your R code in the code chunk provided and your interpretation in the text below it. Aim to write complete sentences — imagine you are explaining your findings to a manager who has not studied statistics.

You have approximately **40 minutes**.
:::

---

# Setup

Run the setup chunk to load packages and data.

```{r}
library(tidyverse)
library(modelsummary)
library(marginaleffects)
library(broom)

firms <- read_csv("data/bisnode-firms.csv")
```

Take two minutes to familiarise yourself with the data:

```{r}
# Explore the data
glimpse(firms)
```

**Question:** What does one row in this dataset represent? What does `exit = 1` mean?

*Your answer:*

---

# Task 1: Fit a Logistic Regression Model

Fit a logistic regression model that predicts firm exit using the following predictors:

- `sales_log` (log of sales revenue)
- `profit_margin` (profit as a share of sales)
- `employees_log` (log of number of employees)
- `age` (firm age in years)

Call your model `model_base`.

::: {.callout-tip}
Use `glm()` with `family = binomial`. The syntax is otherwise the same as `lm()`.
:::

```{r}
# Fit the logistic regression model
# model_base <- ___
```

Inspect the output:

```{r}
# summary(model_base)
```

**Question 1a:** Which predictors are statistically significant at the 5% level?

*Your answer:*

**Question 1b:** Based on the signs of the coefficients, which factors seem to *increase* the probability of exit, and which seem to *decrease* it? Does this make economic sense?

*Your answer:*

---

# Task 2: Extend Your Model

Now add at least **two more predictors** from the dataset to your model. Use your subject-matter intuition: which firm characteristics might predict exit?

Call your extended model `model_ext`.

```{r}
# Look at available variables
# names(firms)
```

```{r}
# Fit the extended model
# model_ext <- ___
```

Compare your two models side-by-side using `modelsummary()`:

```{r}
# YOUR CODE HERE
```

**Question 2a:** Did adding the new predictors improve the model fit (use AIC to compare: lower is better)?

*Your answer:*

**Question 2b:** Did any coefficients change notably when you added the new variables? What might that indicate?

*Your answer:*

---

# Task 3: Interpret Coefficients as Odds Ratios

For your **extended model**, compute and display the odds ratios (with 95% confidence intervals):

```{r}
# Odds ratios
# exp(coef(model_ext))

# With confidence intervals
# exp(confint(model_ext))
```

**Question 3:** Choose **one** predictor from your model and write a complete, businesslike interpretation of its odds ratio. Use the following template:

> "A one-unit increase in [variable] is associated with a multiplication of the odds of firm exit by [odds ratio]. This means the odds of exit [increase/decrease] by [X]%. This effect is [statistically significant / not statistically significant] at the 5% level."

*Your answer:*

---

# Task 4: Predicted Probabilities for Two Firm Profiles

Define two contrasting firm profiles and compute the predicted probability of exit for each. Choose profiles that tell an interesting story.

Fill in the scenario values below:

```{r}
# scenarios <- tibble(
#   label         = c("Profile A: ___", "Profile B: ___"),
#   sales_log     = c(___, ___),         # replace with values
#   employees_log = c(___, ___),
#   profit_margin = c(___, ___),
#   age           = c(___, ___),
#   # add your additional predictors here
# )

# scenarios <- scenarios |>
#   mutate(pred_prob = predict(model_ext, newdata = scenarios, type = "response"))

```

**Question 4:** Describe your two profiles in plain language. Why did you choose these particular values? What story do the predicted probabilities tell?

*Your answer:*

---

# Task 5: Write a Finding Using Inline R Code

Write a short paragraph (3–5 sentences) summarising one key finding from your analysis. Your paragraph must include at least **two numbers embedded as inline R code** (not typed manually).

---

**Your paragraph:**

According to the logistic regression model, ...

*(Continue writing here, embedding inline R code for your numbers)*

---

# Bonus: Average Marginal Effects

::: {.callout-note}
**Optional** — attempt this if you finish the main tasks early.
:::

Compute the average marginal effects (AMEs) for your extended model:

```{r}
# avg_slopes(...)
```

**Bonus question:** Compare the AME for `profit_margin` to the LPM coefficient for `profit_margin` (you may need to fit a quick LPM). Are they similar? What does this tell you about when the LPM might be acceptable?

```{r}
# Fit LPM for comparison
# lpm <- lm(exit ~ sales_log + profit_margin + employees_log + age,
#           data = firms)

# Compare coefficients

```

*Your answer:*

---

# Wrap-up Reflection

Before we debrief, take 2 minutes to note:

1. What was the most conceptually challenging part of this exercise?

2. If you had to explain logistic regression to a colleague with no statistics background, what analogy or example would you use?

*Your answers:*
