---
# ── Document metadata ──────────────────────────────────────────────────────────
title: "CO2 Emissions and Economic Development: A Global Perspective"
subtitle: "A short analytical report"
author: "Prof. Dr. Claudius Gräbner-Radkowitsch"  # change to your name
date: today        # renders as the current date automatically — no need to update
# ── Output formats ─────────────────────────────────────────────────────────────
format:
  html:
    toc: true              # table of contents on the side
    toc-depth: 3           # include headings up to level 3 (###)
    number-sections: true  # numbered headings: 1, 1.1, 1.2, 2, ...
    code-fold: true        # code chunks collapsed by default; reader can expand
    code-tools: true       # Allows users to download source file
  docx:
    toc: true
    number-sections: true
execute: 
  warning: false # Do not show warnings
  message: false # Do not show messages
# ── Bibliography ───────────────────────────────────────────────────────────────
bibliography: references.bib  # path to .bib file (relative to this document)
# ── Recommended project folder structure (keep this for reference) ─────────────
# session_01/
# ├── demo.qmd         ← this file
# ├── fetch_data.R     ← run once to download and save the data
# ├── references.bib   ← bibliography
# ├── data/            ← raw data files go here
# ├── figures/         ← exported figures (if saving separately)
# └── output/          ← rendered reports
---

```{r}
#| label: setup
#| include: false   # this chunk runs but produces no output in the document
here::i_am("content/material/session01/demo.qmd")
library(here)
library(tidyverse)
library(ggrepel)
library(modelsummary)

# Set a consistent ggplot theme for all figures in this document
theme_set(theme_minimal(base_size = 13))
```

## Introduction

Do richer countries pollute more? At first glance, the answer seems obvious: higher
incomes mean more factories, more cars, more flights. But the reality is more nuanced.
Some of the world's largest emitters are middle-income countries with enormous
populations, while several wealthy nations have managed to decouple economic growth
from emissions growth.

In this report, we explore the cross-country relationship between economic output
and CO2 emissions using recent data from the World Bank. The central question is:
**how closely does a country's income level predict its CO2 footprint — and does
population size explain the pattern?**

Understanding this relationship matters for international business: companies
operating across borders face very different regulatory environments and carbon
cost structures depending on where a country sits in the income-emissions space.

## Data and approach

We use panel data covering the period 2020–2024 for all countries for which
complete data are available. The three key variables are:

- **CO2 emissions** (million metric tons CO2 equivalent, AR5 methodology): total
  greenhouse gas emissions attributed to each country
- **GDP** (USD, PPP-adjusted): total economic output, converted to a common
  price level so countries can be compared directly
- **Population**: used to rescale both variables to a per-capita basis and to
  set bubble sizes in the chart

The data were downloaded from the World Bank World Development Indicators using
the `WDI` R package [@WDI2024] and are stored locally in `data/wdi_co2_gdp_pop.csv`.
To reproduce the download, run `fetch_data.R` in this session's folder.

```{r}
#| label: load-data
#| message: false

co2_data <- read_csv(here("content/material/session01/data/wdi_co2_gdp_pop.csv"))
```

For the main visualisation, we average each variable across all available years
per country. This smooths out year-to-year fluctuations and gives a more stable
cross-country picture.

```{r}
#| label: summarise-data

co2_avg <- co2_data |>
  group_by(country, iso2c, iso3c) |>
  summarise(
    across(c(co2_mt, gdp_ppp, pop, gdp_pc_ppp, co2_pc_t), mean),
    years_n = n(),          # how many years contributed to each average
    .groups = "drop"
  ) |>
  filter(pop >= 1e6)        # drop micro-states (< 1 million inhabitants)
```

## Results

### The income–emissions relationship

```{r}
#| label: fig-bubble
#| fig-cap: "GDP per capita and CO2 emissions per capita across countries (2020–2024 average). Bubble size reflects population. Both axes use a log scale."
#| fig-alt: "Bubble chart showing CO2 emissions per capita on the y-axis and GDP per capita on the x-axis, with bubble size proportional to population. A positive association is visible."
#| fig-width: 9
#| fig-height: 6

# Highlight a selection of named countries for orientation
label_countries <- c(
  "United States", "China", "India", "Germany",
  "Brazil", "Nigeria", "Norway", "Saudi Arabia", "Indonesia"
)

ggplot(
  co2_avg,
  aes(x = gdp_pc_ppp, y = co2_pc_t, size = pop, label = country)
) +
  geom_point(alpha = 0.5, colour = "#00395B") +
  geom_smooth(method = "lm", se = FALSE, color = "#69aacd", show.legend = FALSE) +
  geom_text_repel(
    data = co2_avg |> filter(country %in% label_countries),
    size = 3, colour = "#444444", max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_x_log10(labels = scales::label_dollar(scale = 1/1000, suffix = "k")) +
  scale_y_log10(labels = scales::label_comma(suffix = " t")) +
  scale_size_continuous(
    range  = c(1, 18),
    labels = scales::label_comma(scale = 1/1e6, suffix = "M")
  ) +
  labs(
    x    = "GDP per capita (log scale, USD PPP)",
    y    = "CO2 per capita (log scale, metric tons)",
    size = "Population",
    title = "Richer countries tend to emit more CO2 per person",
    subtitle = "But population size shapes total emissions — note China and India"
  )
```

@fig-bubble reveals a clear positive association: countries with higher GDP per
capita also tend to have higher CO2 emissions per person. Both axes use a
**log scale** — this is important when variables span several orders of magnitude,
as both income and emissions do. On a log scale, equal distances represent equal
*percentage* differences, which makes cross-country comparisons more meaningful.

Notice that the relationship is not perfectly tight. Some high-income countries
(e.g. Norway) emit substantially less than their income level would predict —
reflecting a combination of renewable energy, mild climate, and policy choices.
Conversely, some oil-producing economies emit far more than their income alone
would suggest.

The bubble size encodes **population**, which drives a wedge between per-capita
and total emissions. China and India, for example, sit at moderate income and
emissions *per capita*, but their enormous populations make them among the
world's largest total emitters.

### A first regression

```{r}
#| label: tbl-regression
#| tbl-cap: "OLS regression of log CO2 per capita on log GDP per capita (2020–2024 country averages)."

# Both variables already transformed in the data; log-transform here for the model
co2_model <- co2_avg |>
  mutate(
    log_co2_pc  = log(co2_pc_t),
    log_gdp_pc  = log(gdp_pc_ppp)
  )

model <- lm(log_co2_pc ~ log_gdp_pc, data = co2_model)

modelsummary(
  model,
  stars   = TRUE,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Intercept",
    "log_gdp_pc"  = "Log GDP per capita"
  )
)
```

When both variables are log-transformed, the coefficient has a convenient
interpretation: a **1% increase in GDP per capita** is associated with on average a
`r round(coef(model)["log_gdp_pc"], 2)`% change in CO2 per capita, on average.
This is an *elasticity* — a concept we return to in Session 2.

## Conclusion

A simple cross-country regression confirms a strong positive relationship between
income and CO2 emissions per capita. Yet income explains only part of the story:
the residual variation reflects differences in energy mix, industrial structure,
climate policy, and geography. Unpacking these patterns — and distinguishing
correlation from causation — requires the tools we develop across the rest of
this course.

## References
