---
title: "Income and Life Expectancy: A Global Perspective"
author: "Student Name"
date: today
format:
  html:
    toc: true
    number-sections: true
  docx:
    toc: true
    number-sections: true
bibliography: references.bib
---

## Introduction

How does a country's wealth relate to the health of its population? This report
examines the cross-country relationship between income and life expectancy using
data from the `gapminder` package. The relationship between income and life
expectancy was first systematically documented by @Preston1975, who showed that
richer countries tend to have substantially longer life expectancy — a pattern
now known as the *Preston curve*.

Understanding this relationship matters for international business: companies
expanding into lower-income markets face very different consumer health profiles,
healthcare costs, and workforce productivity than those operating in wealthy
economies.

## Data

The analysis uses data from the `gapminder` package [@BekezKezdi2021], which
provides information on GDP per capita, life expectancy, and population for
142 countries observed at five-year intervals from 1952 to 2007. We focus on
the 2007 cross-section — the most recent year in the dataset — to examine the
contemporary income–health relationship.

## Analysis

```{r}
#| label: fig-gapminder
#| fig-cap: "GDP per capita and life expectancy across countries (2007). Bubble size reflects population size."
#| warning: false
#| message: false

library(tidyverse)
library(gapminder)

gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent, size = pop)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "GDP per capita (USD, log scale)",
    y      = "Life expectancy (years)",
    colour = "Continent",
    size   = "Population"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@fig-gapminder shows a clear positive association: countries with higher GDP
per capita tend to have longer life expectancy. The x-axis uses a **log scale**
because GDP per capita varies enormously across countries — on a log scale,
equal distances represent equal *percentage* differences, which makes the
comparison more meaningful.

African countries (red) cluster in the lower-left quadrant — lower income and
shorter life expectancy — while European and Oceanian countries dominate the
upper-right. Asia shows the most heterogeneity: countries like Japan sit near
the top, while several lower-income Asian nations fall much lower. The
relationship is not perfectly tight: some countries have higher life expectancy
than their income level alone would predict, reflecting differences in health
systems, diet, and geography.

```{r}
#| label: fig-gapminder-1952
#| fig-cap: "GDP per capita and life expectancy across countries (1952). Compare with Figure 1 to see how the relationship has changed."
#| warning: false
#| message: false

gapminder |>
  filter(year == 1952) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent, size = pop)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "GDP per capita (USD, log scale)",
    y      = "Life expectancy (years)",
    colour = "Continent",
    size   = "Population"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@fig-gapminder-1952 shows the same relationship in 1952. Comparing the two
figures, several patterns stand out. First, life expectancy has risen
substantially in all regions — the entire cloud of points has shifted upward.
Second, the spread across countries has *narrowed*: in 1952 many countries had
life expectancies below 40 years; by 2007 very few do. Third, the positive
income–health association was already present in 1952, suggesting it is a
persistent structural feature rather than a recent development.

## References
