---
title: "Session 5: Causation vs. Correlation — Live Demo"
subtitle: "Advanced Data Science · Europa-Universität Flensburg"
author: "Claudius Gräbner-Radkowitsch"
date: "2026-05-21"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
  execute-dir: file
---

# Setup

## Business question

> *Is there a gender wage gap among US workers? And what would it take to interpret a regression coefficient as evidence of discrimination?*

## Packages and data

```{r}
library(tidyverse)
library(broom)
library(modelsummary)
library(patchwork)

raw <- read_csv("data/morg-2014-emp.csv", show_col_types = FALSE)
```

The raw file contains `r nrow(raw)` observations with 23 variables from the CPS MORG. We keep only what we need and construct the relevant variables:

```{r}
cps <- raw |>
  filter(
    !is.na(earnwke), earnwke > 0,
    !is.na(uhours),  uhours  > 0,
    !is.na(occ2012), !is.na(grade92),
    !is.na(age), !is.na(sex)
  ) |>
  mutate(
    wage   = earnwke / uhours,
    female = as.integer(sex == 2),
    educ   = grade92,
    occ    = factor(occ2012)
  ) |>
  filter(wage > 0, wage < 500) |>
  select(wage, female, age, educ, occ)

glimpse(cps)
```

```{r}
#| include: false
n_obs    <- nrow(cps)
n_female <- sum(cps$female)
med_m    <- cps |> filter(female == 0) |> pull(wage) |> median() |> round(2)
med_f    <- cps |> filter(female == 1) |> pull(wage) |> median() |> round(2)
raw_gap  <- round((med_m - med_f) / med_m * 100, 1)
```

We have `r n_obs` workers after cleaning. The median hourly wage is \$`r med_m` for men and \$`r med_f` for women — a raw gap of `r raw_gap`%.

---

# The raw picture

## Summary statistics by gender

```{r}
cps |>
  mutate(gender = if_else(female == 1, "Female", "Male")) |>
  group_by(gender) |>
  summarise(
    N            = n(),
    `Median wage`  = median(wage)    |> round(2),
    `Mean wage`    = mean(wage)      |> round(2),
    `SD wage`      = sd(wage)        |> round(2),
    `Median age`   = median(age),
    `Median educ`  = median(educ)
  )
```

## Wage distribution by gender

```{r}
#| label: fig-wage-dist
#| fig-cap: "Log wage distributions by gender. The female distribution sits to the left — but the overlap is large."

cps |>
  mutate(gender = if_else(female == 1, "Female", "Male")) |>
  ggplot(aes(log(wage), fill = gender)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = c("Female" = "tomato", "Male" = "steelblue")) +
  labs(
    x     = "log(hourly wage, USD)",
    y     = "Density",
    fill  = NULL,
    title = "Wage distribution by gender",
    subtitle = paste0("CPS 2014 · n = ", format(n_obs, big.mark = ","))
  ) +
  theme_minimal()
```

There is a clear leftward shift for women, but the distributions overlap substantially. Before concluding this is discrimination, we need to ask: *what else differs between men and women in this sample?*

---

# Three models, three questions

The core claim of this session: every regression answers a specific question. The question changes when you add or remove controls. This is not a matter of some models being "more controlled" — it is a matter of which causal pathway you are estimating.

## Estimating the models

```{r}
m1 <- lm(log(wage) ~ female,                    data = cps)
m2 <- lm(log(wage) ~ female + age + educ,        data = cps)
m3 <- lm(log(wage) ~ female + age + educ + occ,  data = cps)
```

```{r}
#| include: false
gap_m1 <- coef(m1)["female"] |> round(3)
gap_m2 <- coef(m2)["female"] |> round(3)
gap_m3 <- coef(m3)["female"] |> round(3)
```

## The side-by-side comparison

```{r}
#| label: tbl-models
#| tbl-cap: "Three models of the gender wage gap. Outcome: log(hourly wage)."

modelsummary(
  list(
    "Model 1: Raw gap"        = m1,
    "Model 2: + Demographics" = m2,
    "Model 3: + Occupation"   = m3
  ),
  coef_map = c(
    "female" = "Female",
    "age"    = "Age",
    "educ"   = "Education (grade92)"
  ),
  gof_omit = "AIC|BIC|Log|F|RMSE",
  stars    = TRUE,
  notes    = "Occupation dummies (Model 3) omitted from display."
)
```

The female coefficient moves from `r gap_m1` (Model 1) to `r gap_m2` (Model 2) to `r gap_m3` (Model 3). **This does not mean Model 3 is the most correct.** Each model answers a different question.

---

# DAG-based interpretation

## What each model claims

**Model 1 — the raw gap**

$$\log(\text{wage}_i) = \alpha + \beta_1 \cdot \text{female}_i + \varepsilon_i$$

This is a pure descriptive statement: female workers earn, on average, `r abs(gap_m1) * 100`% less per hour in log terms. It makes no causal claim. The coefficient absorbs everything that co-varies with gender: age, education, occupation, unobserved family background, and actual pay discrimination.

**Model 2 — controlling for demographics**

$$\log(\text{wage}_i) = \alpha + \beta_1 \cdot \text{female}_i + \beta_2 \cdot \text{age}_i + \beta_3 \cdot \text{educ}_i + \varepsilon_i$$

Age and education are plausible confounders: they affect wages and are systematically related to gender (the sample shows some difference in educational composition). Adding them closes those back-door paths — but only if age and education are truly confounders and not mediators.

::: {.callout-note}
Are age and education confounders or mediators here? They are treated as confounders — they affect wages independently of gender, and they are not caused by gender. This is defensible if we think of "gender" as a fixed attribute the worker enters the labour market with.
:::

**Model 3 — adding occupation**

$$\log(\text{wage}_i) = \alpha + \beta_1 \cdot \text{female}_i + \beta_2 \cdot \text{age}_i + \beta_3 \cdot \text{educ}_i + \sum_k \gamma_k \cdot \text{occ}_{ik} + \varepsilon_i$$

This is the key model — and the key trap.

If gender discrimination causes occupational sorting (women are pushed toward lower-paying occupations by social and structural forces), then occupation is a **mediator**: it lies *on the causal path* from gender to wage. Controlling for it blocks part of the very effect we are trying to measure.

The coefficient `r gap_m3` is not a more precise estimate of discrimination. It is an estimate of the pay gap **within** occupation groups — the direct channel of discrimination, net of occupational sorting. Whether that narrower quantity is what we want depends entirely on the research question.

## The bad control in plain terms

::: {.callout-important}
**The bad control lesson:** Adding occupation does not make the estimate better. It answers a different question: *"Do women earn less than men in the same occupation, at the same age and education level?"* That is sometimes the right question. But if we want to know the total effect of gender on wages — including the sorting mechanism — Model 3 gives us the wrong answer, not a better one.
:::

```{r}
#| label: fig-gap-shrinkage
#| fig-cap: "The gender coefficient across three models. The shrinkage from Model 2 to Model 3 is not precision — it is a change in the question being answered."

tibble(
  model = factor(
    c("Model 1\nRaw gap", "Model 2\n+ Demographics", "Model 3\n+ Occupation"),
    levels = c("Model 1\nRaw gap", "Model 2\n+ Demographics", "Model 3\n+ Occupation")
  ),
  estimate = c(gap_m1, gap_m2, gap_m3),
  se       = c(
    summary(m1)$coefficients["female", "Std. Error"],
    summary(m2)$coefficients["female", "Std. Error"],
    summary(m3)$coefficients["female", "Std. Error"]
  )
) |>
  ggplot(aes(model, estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray50") +
  geom_pointrange(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    colour = "steelblue", linewidth = 0.8, size = 0.7
  ) +
  labs(
    x     = NULL,
    y     = "Coefficient on Female",
    title = "Gender wage gap across specifications",
    subtitle = "95% confidence intervals shown"
  ) +
  theme_minimal()
```

---

# The identification gap

## Why none of these models proves discrimination

Each model is a conditional association. Even Model 2 — our best effort at controlling for observable confounders — cannot rule out unobserved differences between men and women that drive both wages and are correlated with gender: differences in negotiation behaviour, network access, industry composition, or employer-level pay-setting practices, for example.

An **identification argument** would need to claim: *"The variation in gender I am exploiting is as good as random with respect to wages, conditional on my controls."* No observational regression on this data can make that claim credibly.

::: {.callout-note}
This is not a failure of the data or the technique. It is a property of the question. Gender is not randomly assigned, and the labour market is not an experiment.
:::

## What would help?

Two strategies introduced in upcoming sessions address similar problems:

- **Panel fixed effects (Session 6):** if we had panel data on the same workers over time, we could control for *all* stable unobserved worker-level characteristics. This would eliminate time-invariant confounders but not time-varying ones.

- **Randomised experiments (Session 7):** résumé audit studies — randomly assigning gender-signalling names to identical applications — provide clean causal identification of name-based discrimination. This is a real identification strategy for *one* narrow channel of the gap.

Neither strategy eliminates all confounding for the full gender wage gap. The point is that credibility comes from the *research design*, not from adding more controls.

---

# Summary

| Model | Controls | Question answered | Causal? |
|-------|----------|-------------------|---------|
| 1 | None | Average raw difference in log wages | No — pure association |
| 2 | Age, education | Gap net of demographic composition | Partially — if no unobserved confounders |
| 3 | + Occupation | Gap within occupation, conditional on demographics | Different question — occupation is a mediator |

The key insight: the coefficient shrinks across models not because we are getting closer to the truth, but because we are changing the question. Understanding *which* question each model answers is the skill this session develops.
