---
title: "Session 6: Panel Data and Fixed Effects"
subtitle: "Live Demo & Lecture Notes"
author: "Claudius Gräbner-Radkowitsch"
date: "28 May 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

## Business question

> *Does better management quality lead to better firm performance —
> and does this relationship hold within countries?*

This question matters for managers, investors, and policy-makers alike.
If better-managed firms perform better, investing in management quality
pays off. But answering it carefully requires panel data methods: firms
differ in many stable, hard-to-measure ways (culture, sector, ownership
history), and a naive cross-sectional comparison will confuse management
quality with those background characteristics.

## Packages and data

```{r}
#| label: setup
library(tidyverse)
library(fixest)
library(modelsummary)
```

We use the `wms-management-survey` dataset from Békés & Kézdi (2021),
containing manufacturing firms across 24 countries surveyed between
2004 and 2015. The same firm can appear in multiple survey waves,
giving us the panel structure we need.

```{r}
#| label: load-data
wms <- read_csv("data/wms_da_textbook.csv")

glimpse(wms)
```

The key variables we will use today:

| Variable | Description |
|---|---|
| `firmid` | Unique firm identifier |
| `cty` | Country code |
| `wave` | Survey wave (year) |
| `management` | Management quality score (1–5 scale) |
| `emp_firm` | Number of employees |
| `degree_nm` | Share of non-managers with college degree |
| `degree_m` | Share of managers with college degree |
| `ownership` | Ownership type (founder, family, dispersed, etc.) |

---

# What Is Panel Data?

## Data structure

Our dataset has the classic panel structure: the same units (firms)
observed at multiple time points (survey waves).

```{r}
#| label: panel-structure

# How many firms? How many waves?
wms |>
  summarise(
    n_firms = n_distinct(firmid),
    n_waves = n_distinct(wave),
    n_obs   = n()
  )
```

```{r}
#| label: panel-balance

# Is the panel balanced? How many times is each firm observed?
wms |>
  count(firmid) |>
  count(n, name = "n_firms") |>
  rename(times_observed = n)
```

Most firms appear only once or twice — this is an **unbalanced panel**,
which is typical for survey-based firm data. Fixed effects estimation
handles unbalanced panels without any special adjustments.

## The composite error: motivating fixed effects

A naive approach: regress firm performance (proxied by `management` score
as outcome, or alternatively a performance measure) on firm size and
human capital. But firms differ in many unobserved stable ways —
their founding era, their industry norms, their country's business culture.

Let us make this concrete. Do firms in different countries have
systematically different management scores?

```{r}
#| label: country-variation

wms |>
  group_by(cty) |>
  summarise(mean_mgmt = mean(management, na.rm = TRUE)) |>
  arrange(desc(mean_mgmt)) |>
  ggplot(aes(x = reorder(cty, mean_mgmt), y = mean_mgmt)) +
  geom_col(fill = "#2C5F8A", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Average management score by country",
    x = NULL,
    y = "Mean management score (1–5)"
  ) +
  theme_minimal(base_size = 12)
```

Large cross-country differences exist. If we regress management on
firm size without accounting for country, we are comparing German firms
to Mexican firms — and country differences will contaminate the result.

---

# The Fixed Effects Estimator

## Pooled OLS as a baseline

Let us start with the naive pooled OLS: regress management score on
log firm size and manager education, ignoring any panel structure.

```{r}
#| label: ols-pooled

wms <- wms |>
  mutate(log_emp = log(emp_firm))

m_ols <- feols(management ~ log_emp + degree_m,
               data = wms)

m_ols
```

This estimate mixes together two types of variation:

- **Between firms:** do larger firms have better management on average?
- **Within firms:** when a firm grows, does its management improve?

The pooled OLS coefficient reflects both, dominated by the between
comparison — which is contaminated by unobserved firm and country
characteristics.

## Visualising within vs. between variation

Before running FE, it helps to see the difference intuitively.
Let us pick a handful of countries and plot management vs. firm size,
highlighting the cross-country vs. within-country patterns.

```{r}
#| label: within-between-plot

wms |>
  filter(cty %in% c("US", "DE", "IN", "MX", "GB")) |>
  ggplot(aes(x = log_emp, y = management, colour = cty)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    title = "Management score vs. firm size, selected countries",
    subtitle = "Between-country differences are large; within-country slopes may differ",
    x = "Log(employees)",
    y = "Management score (1–5)",
    colour = "Country"
  ) +
  theme_minimal(base_size = 12)
```

The between-country differences (gaps between the coloured clouds) are
large. Fixed effects removes those gaps and estimates the
**within-country** slope.

## The within-transformation by hand

Before using `feols()`, let us see the within-transformation manually
for one country. The idea: subtract each firm's own mean over time,
leaving only within-firm variation.

```{r}
#| label: within-transform

# For firms with multiple observations: demean within firm
wms_demeaned <- wms |>
  group_by(firmid) |>
  mutate(
    management_dm = management - mean(management, na.rm = TRUE),
    log_emp_dm    = log_emp   - mean(log_emp,   na.rm = TRUE),
    degree_m_dm   = degree_m  - mean(degree_m,  na.rm = TRUE)
  ) |>
  ungroup()

# Run OLS on demeaned data (no intercept — it was removed by demeaning)
m_fe_manual <- feols(management_dm ~ log_emp_dm + degree_m_dm - 1,
                     data = wms_demeaned)

m_fe_manual
```

## Unit FE with `feols()`

`feols()` does the within-transformation automatically and efficiently.
The `|` operator specifies the fixed effects:

```{r}
#| label: unit-fe

m_fe_firm <- feols(management ~ log_emp + degree_m | firmid,
                   vcov = ~firmid,
                   data = wms)

m_fe_firm
```

Notice two things:

1. The coefficient on `log_emp` has likely changed relative to pooled OLS.
   That is the fixed effects doing their job — removing between-firm noise.
2. `firmid` fixed effects are not shown as coefficient rows. They are
   absorbed into unit-specific intercepts.

## Country fixed effects vs. firm fixed effects

It is worth pausing here. We can include FE at different levels:

```{r}
#| label: country-fe

# Country fixed effects — controls for stable country characteristics
m_fe_cty <- feols(management ~ log_emp + degree_m | cty,
                  vcov = ~cty,
                  data = wms)

m_fe_cty
```

Country FE answers: within a country, do larger firms have better management?
Firm FE (below) answers: when a firm grows, does its management improve?

These are different questions with potentially different answers.

---

# Two-Way Fixed Effects

## Adding time fixed effects

Even within countries, there may be year-specific shocks: a global
management consulting boom, a financial crisis that restructures firms,
or changes in survey methodology. Time fixed effects control for anything
affecting all firms in a given year.

```{r}
#| label: twoway-fe

m_fe_2way <- feols(management ~ log_emp + degree_m | firmid + wave,
                   vcov = ~firmid,
                   data = wms)

m_fe_2way
```

The coefficient on `log_emp` is now identified from variation that is
neither explained by which firm it is nor which year it is —
pure within-firm, within-year deviations from both baselines.

## Comparing models side by side

Now we can present the full progression clearly:

```{r}
#| label: model-comparison

modelsummary(
  list(
    "Pooled OLS"  = m_ols,
    "Country FE"  = m_fe_cty,
    "Firm FE"     = m_fe_firm,
    "Two-way FE"  = m_fe_2way
  ),
  gof_map = c("nobs", "r.squared", "r2.within"),
  stars   = TRUE,
  notes   = "Standard errors: pooled OLS uses HC1; FE models clustered at firm level."
)
```

**How to read this table:**

- Moving from Pooled OLS to Country FE: removes cross-country noise
- Moving from Country FE to Firm FE: removes all stable firm-level
  differences — the most important step
- Moving from Firm FE to Two-way FE: additionally removes year-specific
  shocks common to all firms

The within-$R^2$ (row `r2.within`) is the relevant fit statistic for
FE models — it measures fit on demeaned data only.

---

# Interpreting FE Results

## What the coefficient means

In the two-way FE model, $\hat\beta_{log\_emp}$ answers:

> For the *same firm*, when employment increases by 1% (i.e.,
> log-employment increases by 0.01), how much does the management
> score change, holding constant any year-specific shock?

This is a within-firm estimate. We are not comparing Siemens to a
small Mexican manufacturer — we are comparing Siemens to itself in
an earlier or later wave.

## What FE cannot tell us

Fixed effects removes all time-invariant variation.
This means we cannot estimate:

- The effect of `ownership` type (if ownership does not change across waves)
- Country-level effects with firm FE
- Any covariate that is constant within units over time

```{r}
#| label: time-invariant-warning

# Ownership type — does it vary within firms across waves?
wms |>
  group_by(firmid) |>
  summarise(n_ownership_types = n_distinct(ownership)) |>
  count(n_ownership_types)
```

If ownership is mostly constant within firms (which it typically is),
including it alongside firm FE will either drop or produce unreliable estimates.

## Clustered standard errors: why they matter

Errors within the same firm across years are correlated. Let us compare
standard (non-clustered) and clustered standard errors to see the difference:

```{r}
#| label: se-comparison

m_fe_nocluster <- feols(management ~ log_emp + degree_m | firmid + wave,
                        data = wms)   # default: heteroskedasticity-robust

m_fe_cluster <- feols(management ~ log_emp + degree_m | firmid + wave,
                      vcov = ~firmid,
                      data = wms)   # clustered at firm level

modelsummary(
  list("Non-clustered" = m_fe_nocluster, "Clustered (firm)" = m_fe_cluster),
  gof_map = c("nobs", "r2.within"),
  stars   = TRUE,
  notes   = "Two-way FE (firm + wave) in both models."
)
```

Clustered standard errors are typically larger — this is the correct
direction. Non-clustered SEs understate uncertainty by ignoring
within-firm correlation across time.

---

# Quarto Skill: Presenting Panel Results

When you present panel results in a report, three things matter:

**1. Show the progression.** Always include at least pooled OLS and your
main FE specification. This makes the effect of each layer of controls
visible.

**2. Name the fixed effects clearly.** Add a row or note indicating
which FEs are included in each column. `modelsummary` does this
automatically when you use `feols()` — it adds a row labelling the FEs.

**3. Report the within-$R^2$.** Use `gof_map` to select it explicitly:

```{r}
#| label: quarto-skill-demo

modelsummary(
  list(
    "Pooled OLS" = m_ols,
    "Firm FE"    = m_fe_firm,
    "Two-way FE" = m_fe_2way
  ),
  gof_map = c("nobs", "r.squared", "r2.within"),
  stars   = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  notes   = c(
    "Standard errors in parentheses.",
    "FE models: SEs clustered at firm level.",
    "* p<0.1, ** p<0.05, *** p<0.01"
  )
)
```

**Tip:** In a polished report, use a footnote rather than a coefficient
row to communicate which FEs are included. For example: *"Firm and wave
fixed effects included in columns (2) and (3)."*

---

# Summary

Panel data — the same units observed over time — allow us to control
for stable, unobserved unit-level characteristics that would otherwise
bias our estimates.

**Key concepts:**

- **Composite error** $\epsilon_{it} = \eta_i + v_{it}$: the stable
  unit component $\eta_i$ causes OVB when correlated with $x_{it}$
- **Within-transformation:** subtract each unit's time-mean; $\eta_i$
  cancels out; OLS on demeaned data = FE estimator
- **Between vs. within:** two different questions — be deliberate
- **Two-way FE:** adds time fixed effects $\gamma_t$ for common
  period shocks
- **Clustered SEs:** always cluster at the unit level in panel applications
- **Within-$R^2$:** the relevant fit statistic for FE models

**In R:** `fixest::feols(y ~ x | unit_id + time_id, vcov = ~unit_id)`

---

# Further Reading

- Békés, G. & Kézdi, G. (2021). *Data Analysis for Business, Economics,
  and Policy*. Cambridge UP. Ch. 24 (Panel data).
  Available at [gabors-data-analysis.com](https://gabors-data-analysis.com)
- Huntington-Klein, N. (2022). *The Effect: An Introduction to Research
  Design and Causality*. Ch. 16 (Fixed Effects).
  Available at [theeffectbook.net](https://theeffectbook.net)
- Cunningham, S. (2021). *Causal Inference: The Mixtape*. Ch. 8
  (Panel data and fixed effects). Available at
  [mixtape.scunning.com](https://mixtape.scunning.com)
