---
title: "Session 4: Biases and Diagnostics — Live Demo"
subtitle: "Advanced Data Science · Europa-Universität Flensburg"
author: "Claudius Gräbner-Radkowitsch"
date: "2026-05-21"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  execute-dir: file
---

# Setup

## Business question

> *Does the hotel price–distance relationship hold across all European cities, or does it break down in some? And how do we know when to trust our regression output?*

## Packages and data

```{r}
library(tidyverse)
library(broom)
library(modelsummary)
library(lmtest)
library(sandwich)
library(car)
library(performance)
library(see)
library(patchwork)
library(flextable)

hotels <- read_csv("data/hotels_cs.csv", show_col_types = FALSE)
glimpse(hotels)
```

```{r}
#| include: false
n_obs     <- nrow(hotels)
n_cities  <- n_distinct(hotels$city)
med_cheap <- hotels |> filter(city == "Istanbul") |> pull(price) |> median() |> round()
med_exp   <- hotels |> filter(city == "London")   |> pull(price) |> median() |> round()
```

We have a cross-section of `r n_obs` hotels from `r n_cities` European cities (November 2017, complete cases), ranging from Istanbul (median price ~€`r med_cheap`) to London (~€`r med_exp`). The variation across cities shows up not just in average prices but in the *structure* of the data — making this a good setting for exploring what can go wrong in a regression.

```{r}
#| label: tbl-cities
#| tbl-cap: "Overview of the eight cities in the cross-section."

hotels |>
  group_by(city) |>
  summarise(
    N             = n(),
    `Median price (€)`      = median(price)   |> round(1),
    `Median distance (km)`  = median(distance) |> round(2),
    `Median stars`          = median(stars)    |> round(1)
  ) |>
  arrange(`Median price (€)`) |>
  rename(City = city) |>
  flextable() |>
  autofit() |>
  set_caption("Overview of the eight cities in the cross-section.")
```

## Two models we will diagnose

We start with a simple level-level model, then decide whether the log-log specification is better.

```{r}
model_base <- lm(price      ~ distance,                     data = hotels)
model_ext  <- lm(price      ~ distance + stars,             data = hotels)
model_work <- lm(log(price) ~ log(distance + 0.1) + stars,  data = hotels)
```

::: {.callout-note collapse="true"}
## Why `log(distance + 0.1)` and not `log(distance)`?

```{r}
#| echo: false
n_zero <- sum(hotels$distance == 0)
```

`r n_zero` hotels in this cross-section have `distance = 0` — they are recorded as being right at the city centre. `log(0)` is undefined ($-\infty$), so a small positive offset is added before taking the log.

The choice of 0.1 is partly conventional: distances are measured in kilometres, and 0.1 km (100 metres) roughly corresponds to the resolution at which hotel locations are typically geocoded to a city reference point. For the vast majority of hotels — whose distances range from 0.1 to 18 km — the offset has negligible impact.

A good practice is to verify that the results are not sensitive to the choice of constant. For example, re-estimating with `log(distance + 0.05)` or `log(distance + 0.5)` should yield very similar coefficients if the model is well-specified and the offset is small relative to typical distances.
:::

The approach in this session: **build every diagnostic plot by hand first**, so you understand where the numbers come from. Once we have seen the mechanics, we use `performance::check_model()` to produce all panels in one call.

---

# Functional Form

## Starting point: is the level-level specification reasonable?

```{r}
modelsummary(model_ext,
             gof_omit    = "AIC|BIC|Log|F",
             stars       = TRUE,
             coef_rename = c("distance" = "Distance (km)", "stars" = "Stars"))
```

The coefficient on distance is negative (further $\rightarrow$ cheaper), stars is positive (higher rated $\rightarrow$ more expensive). This seems plausible. But should we trust the functional form? The first diagnostic is the **residuals-vs-fitted plot**.

## Building the residuals-vs-fitted plot

`broom::augment()` attaches fitted values (`.fitted`) and residuals (`.resid`) directly to the data frame returned by `lm()`. That is all we need to build the plot:

::: {.callout-tip collapse="true"}
## How to extract fitted values and residuals without `augment()`

`augment()` is convenient, but the same quantities are available directly from the model object:

```r
fitted_vals <- fitted(model_ext)      # equivalent: model_ext$fitted.values
residuals_  <- resid(model_ext)       # equivalent: model_ext$residuals
```

`augment()` is preferred in this workflow because it returns a tidy data frame that includes the original variables alongside `.fitted`, `.resid`, `.hat`, `.std.resid`, and `.cooksd`, making it easy to pipe straight into `ggplot2`.
:::

```{r}
#| label: fig-resid-base
#| fig-cap: "Residuals vs. fitted values for the level-level model. The red smoother curves upward — the functional form is misspecified."

augment(model_ext) |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.12, color = "steelblue", size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(x = "Fitted values (€)", y = "Residuals (€)",
       title = "Level-level: price ~ distance + stars",
       subtitle = "The smoother should be flat — it is not") +
  theme_minimal()
```

The upward curve at high fitted values tells us the model systematically underpredicts expensive hotels. The relationship is not linear in levels.

## Comparing specifications

Let us compare the residual plots for two alternative specifications, shown in @fig-resid-compare (also called Tukey-Anscombe plots).

First, a model with a **quadratic distance term**, which allows the price penalty to change with distance:

$$\text{price}_i = \beta_0 + \beta_1 \cdot \text{distance}_i + \beta_2 \cdot \text{distance}_i^2 + \beta_3 \cdot \text{stars}_i + \varepsilon_i$$

Second, a **log-log specification**, which assumes a constant elasticity relationship between price and distance:

$$\log(\text{price}_i) = \beta_0 + \beta_1 \cdot \log(\text{distance}_i + 0.1) + \beta_2 \cdot \text{stars}_i + \varepsilon_i$$

```{r}
#| label: fig-resid-compare
#| fig-cap: "Residuals vs. fitted for three specifications. The log-log model produces the flattest smoother."
#| fig-width: 11
#| fig-height: 3.8

make_resid <- function(m, title) {
  augment(m) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = 0.12, color = "steelblue", size = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 0.9) +
    labs(x = "Fitted values", y = "Residuals", title = title) +
    theme_minimal(base_size = 11)
}

m_quad <- lm(price ~ distance + I(distance^2) + stars, data = hotels)

make_resid(model_ext,  "Level-level") +
make_resid(m_quad,     "Quadratic") +
make_resid(model_work, "Log-log")
```

The log-log model is clearly best. We adopt it as our working model.

## The RESET test: a formal functional form check

The Ramsey RESET (Regression Equation Specification Error Test) checks whether the functional form of a regression is correctly specified. The key idea: if the model is well-specified, adding polynomial powers of the fitted values ($\hat{y}^2, \hat{y}^3, \ldots$) as additional regressors should not improve the fit — the fitted values already summarise all systematic variation that the model can explain.

- $H_0$: the functional form is correctly specified
- $H_a$: at least one polynomial power of $\hat{y}$ has a non-zero coefficient (something is missing)

A significant test statistic (small p-value) signals misspecification but does not tell you *what* is wrong or how to fix it.

```{r}
resettest(model_ext)
resettest(model_work)
```

The level-level model rejects strongly. The log-log model also rejects, but with a much smaller test statistic.

::: {.callout-note}
**Why does the log-log model still fail?** With `r n_obs` observations, the RESET test has very high statistical power — it can detect even tiny, practically irrelevant deviations from the specified form. The remaining systematic structure likely reflects city-level price heterogeneity that no two-variable regression can absorb. This is a preview of the omitted variable problem we address at the end of this session.
:::

::: {.callout-tip}
A useful workflow: RESET first to flag a problem, residual plot to diagnose it, revised model to fix it.
:::

---

# Heteroskedasticity

## Building the scale-location plot

The scale-location plot shows $\sqrt{|\tilde{e}_i|}$ on the y-axis, where $\tilde{e}_i$ are the *standardised* residuals. If error variance is constant, this band should be flat and horizontal. We build it manually from `augment()`:

::: {.callout-tip collapse="true"}
## Computing these values without `augment()`

The standardised residuals are available directly via `rstandard()`:

```r
std_resid <- rstandard(model_work)    # standardised residuals: e_i / (sigma * sqrt(1 - h_ii))
sqrt_abs  <- sqrt(abs(std_resid))
fitted_v  <- fitted(model_work)
```

`rstandard()` divides each residual by $\hat{\sigma}\sqrt{1 - h_{ii}}$, where $h_{ii}$ is the leverage of observation $i$. This scaling makes residuals from high-leverage observations comparable to the rest.
:::

```{r}
#| label: fig-scaleloc-manual
#| fig-cap: "Scale-location plot: the upward slope reveals heteroskedasticity in the log-log model."

augment(model_work) |>
  ggplot(aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.12, color = "steelblue", size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(
    x = "Fitted values (log scale)",
    y = "√|Standardised residuals|",
    title = "Scale-location: log(price) ~ log(distance) + stars",
    subtitle = "A flat horizontal smoother = homoskedastic"
  ) +
  theme_minimal()
```

The smoother rises toward higher fitted values — hotels in expensive cities show more price dispersion than the model can absorb. OLS estimates are still unbiased, but standard errors are unreliable.

## The Breusch-Pagan test

The Breusch-Pagan test formalises what the scale-location plot showed visually. If errors are homoskedastic, the squared residuals $\hat{u}_i^2$ should not be systematically related to the regressors — knowing $X$ should not help predict how large the error will be. The test regresses $\hat{u}_i^2$ on the fitted values (or on the original regressors, depending on the variant) and uses a $\chi^2$-test to check whether this auxiliary regression has any explanatory power:

- $H_0$: homoskedasticity — $\text{Var}(u_i \mid X_i) = \sigma^2$ (constant)
- $H_a$: heteroskedasticity — $\text{Var}(u_i \mid X_i) = h(X_i)$ for some function $h(\cdot)$

Under $H_0$, the test statistic follows a $\chi^2$ distribution with degrees of freedom equal to the number of variables in the auxiliary regression.

```{r}
bptest(model_work)
```

The tiny p-value confirms what the plot showed: the error variance is not constant. We reject homoskedasticity.

## Robust standard errors

The fix is to replace the classical standard errors with heteroskedasticity-consistent (HC) ones. The estimates do not change — only the expressed uncertainty.

```{r}
coeftest(model_work, vcov = vcovHC(model_work, type = "HC1"))
```

For reporting, compare both in a single table:

```{r}
#| label: tbl-robust
#| tbl-cap: "Classical vs. HC1 robust SEs. Estimates identical; inference differs."

modelsummary(
  list("Classical SEs" = model_work, "HC1 robust SEs" = model_work),
  vcov        = list("classical", "HC1"),
  stars       = TRUE,
  gof_omit    = "AIC|BIC|Log|F|RMSE",
  coef_rename = c("log(distance + 0.1)" = "log(distance)", "stars" = "Stars")
)
```

::: {.callout-note}
With cross-sectional data, use robust SEs by default. The additional cost is zero.
:::

---

# Multicollinearity

## VIF for the working model

The Variance Inflation Factor (VIF) for predictor $j$ measures how much collinearity with the other predictors inflates the variance of $\hat{\beta}_j$. It is computed by regressing $x_j$ on all other predictors and using the resulting $R^2_j$:

$$\text{VIF}_j = \frac{1}{1 - R^2_j}$$

A VIF of 1 means no inflation at all; a VIF of 5 means the standard error of $\hat{\beta}_j$ is $\sqrt{5} \approx 2.2$ times larger than it would be without collinearity; a VIF of 10 means roughly a 3-fold inflation. Values above 5 warrant attention; values above 10 are serious.

```{r}
vif(model_work)
```

Both VIFs are close to 1 — no concern. The two predictors are measuring genuinely different things.

## What problematic VIF looks like

A researcher might wonder whether to use `stars` in raw form or log form. Including *both* creates severe collinearity:

```{r}
model_collinear <- lm(log(price) ~ log(distance + 0.1) + stars + log(stars), data = hotels)
vif(model_collinear)
```

VIF ≈ 20 for `stars` and `log(stars)` — a variable and its monotonic transformation are near-identical. The model cannot separate their effects and the standard errors become meaningless. The remedy is straightforward: pick one form and stick with it.

---

# Influential Observations

## Building the leverage-residuals plot

We need three quantities from `augment()`:

- `.hat`: the *leverage* of each observation — how far it sits in predictor space ($h_{ii}$ from the hat matrix)
- `.std.resid`: the *standardised residual* — how unusual the outcome is given the predictors
- `.cooksd`: *Cook's distance* — combines both into a single influence measure

```{r}
hotels_aug     <- augment(model_work) |> bind_cols(hotels |> select(hotel_id, city))
threshold_cook <- 4 / nrow(hotels)
p_params       <- length(coef(model_work))   # number of estimated parameters
```

The straight horizontal lines at ±2 you might see in basic plots are a simplification — they only consider residual size and ignore leverage. An observation with high leverage but a small residual can still be highly influential. The correct visualisation uses **Cook's distance contour curves**, which show equal influence in the joint (leverage, residual) space. The contour at threshold $D^*$ is:

$$|r_i| = \sqrt{D^* \cdot p \cdot \frac{1-h_{ii}}{h_{ii}}}$$

where $p$ is the number of estimated parameters and $h_{ii}$ is the leverage. We add these curves to the manual plot using `threshold_cook = 4/n`.

The y-axis is clipped with `coord_cartesian()` to suppress a handful of extreme standardised residuals — this clips the *display* without dropping any observations from the model.

```{r}
#| label: fig-leverage-manual
#| fig-cap: "Leverage vs. standardised residuals with Cook's distance contour at the 4/n threshold. Red points exceed the threshold. Y-axis clipped at the 99.9th percentile of |std.resid| for readability."

h_seq <- seq(1e-4, max(hotels_aug$.hat) * 1.05, length.out = 500)

contour_df <- tibble(
  h     = h_seq,
  upper =  sqrt(threshold_cook * p_params * (1 - h_seq) / h_seq),
  lower = -sqrt(threshold_cook * p_params * (1 - h_seq) / h_seq)
)

y_lim <- quantile(abs(hotels_aug$.std.resid), 0.999) * 1.1

hotels_aug |>
  ggplot(aes(.hat, .std.resid,
             size  = .cooksd,
             color = .cooksd > threshold_cook)) +
  geom_point(alpha = 0.5) +
  geom_line(data = contour_df, aes(x = h, y = upper),
            inherit.aes = FALSE, color = "tomato", linetype = "dashed", linewidth = 0.7) +
  geom_line(data = contour_df, aes(x = h, y = lower),
            inherit.aes = FALSE, color = "tomato", linetype = "dashed", linewidth = 0.7) +
  coord_cartesian(ylim = c(-y_lim, y_lim)) +
  scale_color_manual(
    values = c("steelblue", "tomato"),
    labels = c("Below threshold", "Cook's D > 4/n"),
    name   = NULL
  ) +
  scale_size_continuous(guide = "none") +
  labs(x = "Leverage (h_ii)", y = "Standardised residuals",
       title    = "Influential observations",
       subtitle = "Dashed curves = Cook's D contour at 4/n") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Who are the most influential hotels?

```{r}
#| label: tbl-influential
#| tbl-cap: "Ten most influential hotels by Cook's distance."

hotels_aug |>
  arrange(desc(.cooksd)) |>
  filter(.cooksd > threshold_cook) |>
  select(city, .cooksd, .hat, .std.resid) |>
  head(10) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  rename(City = city, `Cook's D` = .cooksd,
         Leverage = .hat, `Std. residual` = .std.resid) |>
  flextable() |>
  autofit()
```

## Sensitivity check

```{r}
#| label: tbl-infl-sensitivity
#| tbl-cap: "Model with and without influential observations. The core results are robust."

influential_ids <- hotels_aug |> filter(.cooksd > threshold_cook) |> pull(hotel_id)

model_no_infl <- hotels |>
  filter(!hotel_id %in% influential_ids) |>
  lm(log(price) ~ log(distance + 0.1) + stars, data = _)

modelsummary(
  list("Full sample" = model_work, "Excl. influential" = model_no_infl),
  stars       = TRUE,
  gof_omit    = "AIC|BIC|Log|F|RMSE",
  coef_rename = c("log(distance + 0.1)" = "log(distance)", "stars" = "Stars")
)
```

The coefficients are stable — the influential observations are a concern worth noting but do not overturn the conclusions.

---

# Normality of Residuals

## Building the Q-Q plot

The Q-Q (quantile-quantile) plot compares the empirical distribution of the standardised residuals against the theoretical distribution assumed by OLS — the standard normal. If residuals are normally distributed, their quantiles will align with the theoretical normal quantiles and the points will fall on or near the diagonal.

Deviations from the diagonal in the tails indicate that the residual distribution is heavier- or lighter-tailed than a normal. A pronounced S-shape (common with right-skewed data) means more extreme residuals than the normal distribution predicts.

**Important caveat:** normality of residuals is required for *exact* finite-sample t- and F-tests, but with moderate-to-large $n$, the Central Limit Theorem ensures approximate normality of $\hat{\beta}$ regardless of the residual distribution. For our dataset with `r n_obs` observations, the Q-Q plot is informative but rarely a primary concern.

```{r}
#| label: fig-qq-manual
#| fig-cap: "Q-Q plot: standardised residuals vs. theoretical normal quantiles."

hotels_aug |>
  ggplot(aes(sample = .std.resid)) +
  stat_qq(alpha = 0.15, color = "steelblue", size = 0.7) +
  stat_qq_line(color = "tomato", linewidth = 1) +
  labs(x = "Theoretical normal quantiles",
       y = "Sample quantiles of std. residuals",
       title = "Q-Q plot of standardised residuals") +
  theme_minimal()
```

The tails deviate slightly from the diagonal — heavier than a perfect normal. With $n = `r n_obs`$, the Central Limit Theorem ensures this does not affect our inference.

## The Shapiro-Wilk trap

The Shapiro-Wilk test assesses whether a sample could plausibly have been drawn from a normal distribution. It is based on the correlation between the sample quantiles and the quantiles expected under normality — a value close to 1 indicates normality, and a significant p-value suggests a departure. Because R's `shapiro.test()` is limited to a maximum of 5,000 observations, we draw a random subsample from the residuals rather than testing all `r n_obs` at once.

```{r}
set.seed(42)
shapiro.test(sample(resid(model_work), 5000))
```

p $\ll$ 0.001. With `r n_obs` observations, Shapiro-Wilk has enough statistical power to detect even trivially small departures from normality that have no practical consequence for inference — the large $n$ makes the test hypersensitive. This result is not alarming; use the Q-Q plot for visual assessment instead.

---

# The Posterior Predictive Check

The PPC is the one diagnostic that checks the model's *distributional* assumptions rather than its residuals. It simulates new response values from the fitted model many times and asks: does the distribution of the simulated data resemble the observed data?

## model_base: a distributional failure

```{r}
#| label: fig-ppc-base
#| fig-cap: "PPC for the level-level model. Simulated prices spread far beyond the observed range."

plot(check_predictions(model_base)) +
  labs(title    = "Level-level: price ~ distance + stars",
       subtitle = "Residual skewness = 7.8 — simulated data spreads to €3,500+") +
  theme_minimal()
```

The green (simulated) curves are nearly flat across a range the observed data never occupies. The level-level model generates plausible prices at the wrong scale.

## model_work: much better

```{r}
#| label: fig-ppc-work
#| fig-cap: "PPC for the log-log model. The distributions now broadly align."

plot(check_predictions(model_work)) +
  labs(title    = "Log-log: log(price) ~ log(distance) + stars",
       subtitle = "Residual skewness ≈ 0 — distributions align") +
  theme_minimal()
```

The PPC caught what the residual plot could not show as clearly: the level-level model's assumed data-generating process is fundamentally incompatible with the right-skewed distribution of hotel prices.

---

# Putting It Together: `performance::check_model()`

We have now built each diagnostic plot by hand. `performance::check_model()` produces all of them in one call. Now that you understand what each panel shows, the automated output is much more informative:

```{r}
#| label: fig-checkmodel-final
#| fig-cap: "Complete diagnostic dashboard for the log-log working model."
#| fig-width: 10
#| fig-height: 7

performance::check_model(model_work)
```

| Panel | What it shows | Our finding |
|---|---|---|
| Posterior predictive check | Observed vs. simulated distribution of $y$ | Distributions align — log transformation fixed the failure |
| Linearity | Residuals vs. fitted | Largely flat — log-log is the right form |
| Homogeneity of variance | Scale-location | Remaining fan shape → use robust SEs |
| Influential observations | Cook's D + leverage | Small cluster above threshold; results robust |
| Collinearity | VIF bar chart | Both predictors < 2 — no concern |
| Normality | Q-Q plot | Near-normal; CLT protects inference at $n \approx$ `r n_obs` |

::: {.callout-note  collapse="true"}
## Why does `check_model()` look so different?

You might have noticed that the plot on influential observations provided by `performance::check_model()` looks different to our manual @fig-leverage-manual.
The reasons is as follows: `performance::check_model()` draws contours at a **fixed** Cook's distance threshold (around $D^* = 0.5$–$0.7$), which was designed for small-sample regression. For our dataset with $n = `r nrow(hotels)`$, this threshold is enormous relative to the actual Cook's distances. Plugging into the contour formula at a typical leverage of $h \approx 0.0005$ and $D^* = 0.7$:

$$|r_i| = \sqrt{0.7 \times 3 \times 0.9995 \,/\, 0.0005} \approx 65$$

This pushes the y-axis to ±150, compressing all actual data points near zero and making the panel essentially unreadable. The $4/n$ rule scales with sample size and is the appropriate threshold for large $n$: it treats an observation as influential if removing it changes the fitted values by the equivalent of one standard error on average. In effect, our manual plot using $4/n$ is more informative for this dataset than the `check_model()` panel.
:::
---

# What Diagnostics Cannot Detect

## The limits of residual-based analysis

Every panel above examines properties of the *residuals* — the part of $y$ the model did not explain. The residuals are defined relative to the model we already estimated: if an important variable is missing, the bias is baked into $\hat{\beta}$ before we ever look at the residuals.

This is a general property of *all* routes to endogeneity ($E[u \mid X] \neq 0$):

- **Omitted variable bias** — the confounder's effect is absorbed into $\hat{\beta}$; residuals show no trace
- **Measurement error in regressors** — attenuation bias hides behind well-behaved residuals
- **Simultaneity / reverse causality** — the endogenous regressor leaves no obvious pattern in $\hat{u}$

All three share the same logic: the bias lives *inside* $\hat{\beta}$, not in $\hat{u}$. In what follows we focus on **omitted variable bias** as the most common case in business and economics research.

## Demonstrating OVB: does adding city matter?

```{r}
#| label: tbl-ovb
#| tbl-cap: "Without and with city controls. The distance coefficient shifts substantially."

m_no_city   <- lm(log(price) ~ log(distance + 0.1) + stars,        data = hotels)
m_with_city <- lm(log(price) ~ log(distance + 0.1) + stars + city,  data = hotels)

modelsummary(
  list("Without city" = m_no_city, "With city controls" = m_with_city),
  coef_omit   = "city",
  stars       = TRUE,
  gof_omit    = "AIC|BIC|Log|F|RMSE",
  coef_rename = c("log(distance + 0.1)" = "log(distance)", "stars" = "Stars"),
  add_rows    = tribble(
    ~term,     ~`Without city`, ~`With city controls`,
    "City FE", "No",            "Yes"
  )
)
```

The distance coefficient changes substantially. Why? Expensive cities (Paris, London) are also more compact — hotels sit closer to the centre. Without controlling for city, the model confounds "expensive" with "central" and attributes the price difference to distance.

## The residuals from the biased model look fine

```{r}
#| label: fig-ovb-residuals
#| fig-cap: "Three diagnostic panels for the model omitting city. All pass — yet the distance coefficient is biased."
#| fig-height: 9

p1 <- augment(m_no_city) |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.12, color = "steelblue", size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(x = "Fitted values", y = "Residuals",
       title = "Linearity (residuals vs. fitted)") +
  theme_minimal()

p2 <- augment(m_no_city) |>
  ggplot(aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.12, color = "steelblue", size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(x = "Fitted values", y = "√|Standardised residuals|",
       title = "Homogeneity of variance (scale-location)") +
  theme_minimal()

p3 <- augment(m_no_city) |>
  ggplot(aes(sample = .std.resid)) +
  stat_qq(alpha = 0.15, color = "steelblue", size = 0.7) +
  stat_qq_line(color = "tomato", linewidth = 1) +
  labs(x = "Theoretical quantiles", y = "Sample quantiles",
       title = "Normality of residuals (Q-Q)") +
  theme_minimal()

p1 / p2 / p3
```

Clean residuals, flat smoother, near-normal Q-Q. The bias is invisible because it has been absorbed into $\hat{\beta}_\text{distance}$. This is why OVB — and the other routes to endogeneity — require theoretical reasoning and better research design, not better diagnostics.

## The direction of bias

$$\text{sign(bias)} = \text{sign}(\text{effect of city on price}) \times \text{sign}(\text{corr(city, distance)})$$

- City → price: positive (Paris > Budapest)
- City → distance: negative (expensive cities are compact; hotels are more central)
- Bias on $\hat{\beta}_\text{distance}$: negative → the estimated distance effect is *more negative* than it should be

---

# Summary: the diagnostic checklist

For every regression you report, work through these six questions:

1. **Residuals vs. fitted** — is the functional form right? (`augment()` + ggplot2, or `check_model()` linearity panel)
2. **Scale-location** — is variance constant? If not, use robust SEs (`bptest()`, then `vcovHC()`)
3. **Cook's D / leverage** — are a handful of observations driving your results? (augment + ggplot2, or `check_model()` panel 4)
4. **VIF** — are predictors too correlated? (`car::vif()`, or `check_model()` panel 5)
5. **Q-Q plot** — are there serious distributional departures? (augment + ggplot2, or `check_model()` panel 6)
6. **Posterior predictive check** — does the model generate plausible data? (`check_predictions()`, or `check_model()` panel 1)
7. **Think** — what variable did you leave out, and does measurement error or reverse causality apply? No function answers this.
