---
title: "Exercise: What drives hotel prices in Vienna?"
subtitle: "Session 2 — Multiple Regression: Going Beyond the Basics"
author: "Your Name (Student ID)"
date: "09 April 2026"
format:
  html:
    code-tools: true
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(modelsummary)
library(flextable)
```

**Business Question**

You work as an analyst for a hotel booking platform. Your manager asks:

> *"We know hotels further from the city centre tend to be cheaper. But does that penalty look the same for a budget three-star as for a luxury five-star property? And what else explains price differences once we account for location?"*

In this exercise you will use multiple regression — including a categorical variable and an interaction term — to answer this question.

# Data

**Source:** Békés & Kézdi (2021), *Data Analysis for Business, Economics, and Policy*. Hotels in Vienna for one date.

Download the dataset from OSF: <https://osf.io/4e6d8/files/y6jvb/>

The file you need is `hotels-vienna.csv`. Place it in the `data/` subfolder next to this document.

```{r}
#| label: load-data
#| message: false

hotels <- read_csv("data/hotels-vienna.csv")  |> filter(distance>0)
```

## Explore the data

```{r}
#| label: eda

glimpse(hotels)
```

**Task 1.1** In 2–3 sentences, describe what one observation in this dataset represents. List the variables you plan to use and state whether each is numeric or categorical.

**Task 1.2** Produce simple summary statistics (mean, standard deviation, min, max) for `price`, `distance`, and `stars`. Comment briefly: what is the range of prices? How far from the centre do hotels typically sit?

```{r}
#| label: tbl-descriptives
#| tbl-cap: "Descriptive statistics: Vienna hotels"

# Hint: compute summary statistics with summarise(), build a small data frame
# with one row per variable, then format it with flextable() — the same
# workflow you used in Task 1.
# Give the columns readable names: "Variable", "Mean", "Std. Dev.", "Min", "Max".

# YOUR CODE HERE
```

# Data Preparation

For a focused analysis we restrict the sample to hotels with 3, 4, or 5 stars and drop any observations with missing prices or distances.

```{r}
#| label: data-prep

hotels_clean <- hotels |>
  filter(stars %in% c(3, 3.5, 4, 4.5, 5)) |>
  filter(!is.na(price), !is.na(distance)) |>
  mutate(
    # Create a three-category star variable — budget, mid-range, luxury
    star_cat = case_when(
      stars <= 3   ~ "3-star",
      stars <= 4   ~ "4-star",
      stars >= 4.5 ~ "5-star"
    ),
    # Set 3-star as the reference category
    star_cat = factor(star_cat, levels = c("3-star", "4-star", "5-star"))
  )
```

**Task 2.1** How many observations remain after filtering? How many hotels fall in each `star_cat` group? Use `count()` to find out and state the result in one sentence.

```{r}
#| label: counts

# YOUR CODE HERE
```

## Visualise the key relationship

**Task 2.2** Create a scatter plot of `price` (y-axis) against `distance` (x-axis). If necessary, transform the variables. Colour the points by `star_cat` and add a separate linear regression line for each category using `geom_smooth(method = "lm", se = FALSE)`. Give the plot informative axis labels and a title.

In 2–3 sentences: do the regression lines run roughly parallel, or do they differ in slope? What does this suggest about the relationship between distance, star category, and price?

```{r}
#| label: fig-scatter
#| fig-cap: "Hotel price vs. distance from city centre, Vienna"

# YOUR CODE HERE
```

# Regression Models

## Estimating the models

Based on the scatter plot, use log-transformed price and distance in all models. In a log-log model, coefficients are elasticities: a 1% increase in distance is associated with a β% change in price.

- **Model 1:** Just regress the log of `price` on the log of `distance`
- **Model 2:** Add `stars` as additional explanatory variable
- **Model 3:** Use `star_cat` instead of `stars` additional explanatory variable
- **Model 4:** Extend model 3 to include interaction effects

```{r}
#| label: model1

# m1 <- lm(log(price) ~ ..., data = hotels_clean)
```

```{r}
#| label: model2

# m2 <- lm(..., data = hotels_clean)
```

```{r}
#| label: model3

# m3 <- lm(..., data = hotels_clean)
```

```{r}
#| label: model4

# m4 <- lm(..., data = hotels_clean)
```

## Model Comparison

Produce a single `modelsummary` table containing all four models. Use `coef_map` to rename the row labels, and add a `notes` argument describing the sample and transformations.

```{r}
#| label: tbl-models
#| tbl-cap: "Drivers of hotel prices in Vienna: regression results (log-log)"

coef_labels <- c(
  "(Intercept)"                  = "Constant",
  "log(distance)"                = "log(Distance from centre)",
  "stars"                        = "Star rating (continuous)",
  "star_cat4-star"               = "4-star hotel",
  "star_cat5-star"               = "5-star hotel",
  "log(distance):star_cat4-star" = "log(Distance) × 4-star",
  "log(distance):star_cat5-star" = "log(Distance) × 5-star"
)

# modelsummary(
#   list(
#       ADD HERE
#   ),
#   coef_map = coef_labels,
#   gof_map  = c("nobs", "r.squared", "adj.r.squared"),
#   stars    = TRUE,
#   notes    = ""
# )
```

## Interpretation

Use @tbl-models to answer the following questions.

**Task 3.1** What does the `log(distance)` coefficient in column (1) tell you? Why is a log-log specification more appropriate here than a linear model?

**Task 3.2** In column (2), interpret the `stars` coefficient. What does this model assume about the price step from 3- to 4-star compared with 4- to 5-star? Is that assumption plausible?

**Task 3.3** Answer the following using column (3):

1. What is the reference category?
2. Interpret the `4-star hotel` coefficient. *(Hint: in a log-price model, the percentage premium is `(exp(β) − 1) × 100`.)*
3. How does model (3) differ from model (2) in what it allows the data to tell you?

**Task 3.4** Answer the following using column (4):

1. Interpret the `log(Distance) × 4-star` coefficient. What does it say about how the distance penalty differs for 4-star hotels?
2. What is the total distance elasticity for a 5-star hotel? Show the calculation.
3. Does this match what you saw in the scatter plot?

**Task 3.5** Looking at the fit statistics at the bottom of the table:

1. Does $R^2$ increase as you add more variables? What does the adjusted $R^2$ tell you that the plain $R^2$ does not?
2. Does the `log(distance)` coefficient change substantially between columns (1), (3), and (4)? What does stability or instability suggest?

# Communication

**Task 4** Write a short paragraph (80–120 words) for your manager summarising the findings. Do not use statistical jargon — no p-values, no "coefficients". Focus on the business insight: What drives hotel prices in Vienna? Does being far from the centre hurt all hotels equally? What practical implication does this have for a hotel trying to set its prices?
